Network based multifactorial modelling microRNA-target interaction.

### 2019-09-10 article.Rmd
news to article

---
title: "Network based multifactorial modelling microRNA-target interaction."
author:
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document: default
  pdf_document:
    keep_tex: yes
  html_document: default
documentclass: article
link-citations: yes
cls: NAR.cls
template: NAR-sample.tex
bibliography: ref.bib
---

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other targets' expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from [TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga)(#TODO). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#FIX name of database and its citation) answer->suplementary datasets of [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, expression level change of common target (G4) in network has been observed different from G1 that has the same expression value with G4 target at initial conditions because more than one miRNA repress this target (\autoref{fig1}d) (#FIX last sentence needs some clarification, after "but cooperative...").

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), the changes of other gene expressions have observed more prominent [][#TODO]. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes. [][#TODO2]

@@

[#Add]Similar with this approach, Silveira et al. have developed miRmapper package for using with R programme [@da2018mirmapper]. Researchers have utilized the adjacency matrix to associate miRNAs using differentially expressed genes and found similarities of miRNAs with this approach. Authors who utilized the approach in dataset of bladder cancer cell line have identified importance of miRNAs in network, and miRNAs which work synergistically.

[#Add] Today, there are many of tools that provide predicted or experimentally validated miRNA:target interaction dataset. The datasets that contain weak, proteomic analysis like western blot, and strong, crosslinking and immunoprecipitation CLiP, experimentally evidence are more reliable sources than prediction datasets. While in experimentally weak supported datasets it is not known that indirect or direct target of miRNAs in cell, high-throughput methods ensure handling of bounded miRNA:target pairs. For this reason we have preferred to utilize high-throughput datasets in our application. We collected datasets that depend on chimeric [@helwak_mapping_2013; @moore_mirnatarget_2015] reading of miRNA: target pairs from among these datasets because they contain exact complementary seed regions. Although these experimental sources include exact binding sites, they do not provides which miRNA is functional or not [@liu2019prediction]. We have used the features of miRNA:target pairs and scaled these parameters inside groups that is consisted according to ,ceRNA (Competing Endogenous RNA), genes targeted by the same miRNA. *In scaling process, we have considered previous studies. BURAYA O ÇALIŞMALARI ATIFLA*

>>>>

---
title: "Network based multifactorial modelling microRNA-target interaction."
author:
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
  html_document: default
documentclass: article
link-citations: yes
cls: NAR.cls
template: NAR-sample.tex
bibliography: ref.bib
---

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other targets' expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common targeting miRNA of genes and target of miRNAs, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from The Cancer Genome Atlas [TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga)(#TODO). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#FIX name of database and its citation) answer->suplementary datasets of [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, expression level change of common target (G4) in network has been observed different from G1 that has the same expression value with G4 target at initial conditions because more than one miRNA repress this target (\autoref{fig1}d) (#FIX last sentence needs some clarification, after "but cooperative...").

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), the changes of other gene expressions have observed more prominent [][#TODO]. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes. [][#TODO2]

@@

[#Add]More recently, Silveira et al. have developed miRmapper package for using with R programme [@da2018mirmapper]. Researchers have utilized the adjacency matrix to associate miRNAs using differentially expressed genes and found similarities of miRNAs with this approach. Authors who utilized the approach in dataset of bladder cancer cell line have identified significant miRNAs in network, and miRNAs which work synergistically.

[#Add] Today, there are many of tools that provide predicted or experimentally validated miRNA:target interaction dataset. The datasets that contain weak, proteomic analysis like western blot, and strong, crosslinking and immunoprecipitation CLiP, experimentally evidence are more reliable sources than prediction datasets. While in experimentally weak supported datasets it is not known that indirect or direct target of miRNAs in cell, high-throughput methods ensure handling of AGO (Argounate Protein) interacted miRNAs. For this reason we have preferred to utilize high-throughput datasets in our application. Especially, we collected datasets that depend on chimeric [@helwak_mapping_2013; @moore_mirnatarget_2015] reading of miRNA: target pairs from among these datasets because they contain exact complementary seed regions.

[#Add] Although these experimental sources include exact binding sites, they do not provides which miRNA is functional or not [@liu2019prediction]. We have used the features of bound miRNA:target pairs and scaled these parameters inside groups that is consisted according to, ceRNA (Competing Endogenous RNA), genes targeted by the same miRNA. In scaling process, we specified the interaction factors as numeric values in accordance with previous studies. Energy values in miRNA:target pairs are represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015] which are utilized in this study. On the other hand, we have specified the other interaction factors, seed type and location of binding region on the target, as numeric values based on the previous studies.[@grimson_microrna_2007] have compared the seed types' effect on target repression with few miRNA had canonical seed pairing in their study. Additionally, [@bartel_micrornas:_2009] and [@betel2010comprehensive] have studied on functional and non-functional seed interactions. Based on results of these studies we have arranged seed types of miRNA:target interactions as numeric values. We also have redefined location of binding region on the target as numeric values, based on studies of [@hausser_analysis_2013] and [@helwak_mapping_2013]. With this process, we have handled this entegrated dataset in context of competitor behaviours and functionality of interactions.

[#Add]The main advantage of the method, ceRNAnetsim package, is providing various functions which contain optionally arguments. So, users can run the functions in accordance with their own data. For example, the functions can work without interaction factors or with a simple interaction factor that is specified by user. Furthermore, it can also be operated according to a different parameter(s) provided by the user.In this study, we have preferred to run model, shown in *Multifactorial calculations in miRNA:target network* section.
